### 2022-06-08 manuscript/hiv-gwas.tex
Update on Overleaf.

\begin{abstract}
	Infectious diseases are particularly challenging for genome-wide association studies (GWAS) because genetic effects from two organisms (pathogen and host) may significantly influence a trait of interest. Previous GWAS have successfully identified several human genetic variants associated with HIV-1 set point viral load (spVL), among other important infectious disease traits. However, these GWAS do not account for potentially confounding or extraneous pathogen effects that are heritable from donor to recipient in transmission chains.  We propose a new method to consider the full genome of each patient's infecting pathogen strain, remove strain-specific effects on a trait based on the pathogen phylogeny, and thus better estimate the effect of human genetic variants on infectious disease traits. In simulations, we show our method can increase GWAS power to detect truly associated host variants when pathogen effects are highly heritable, with strong phylogenetic correlations. When we apply our method to HIV-1 subtype B data from the Swiss HIV Cohort Study, we recover slightly weaker but qualitatively similar signals of association between spVL and human genetic variants in the \emph{CCR5} and major histocompatibility complex (MHC) gene regions compared to standard GWAS. Our simulation study confirms that based on the estimated heritability and selection parameters for HIV-1 subtype B spVL, standard GWAS are robust to pathogen effects. Our framework may improve GWAS for other diseases if pathogen effects are even more phylogenetically correlated amongst individuals in a cohort.
\end{abstract}

@@

% How do we measure the extent to which genetic factors determine spVL
To understand HIV pathogenicity, it is important to understand to what extent spVL is determined by host genetic factors \citep{Bartha2013, Dalmasso2008, Fellay2007, Pereyra2010, Fellay2009, Pelak2010, VanManen2011, McLaren2012}. Heritability is a key measure of how genetically-determined a trait is. Here we distinguish between two different heritability measures that are used in different contexts in the study of infectious diseases. Broad-sense heritability $H^2$ measures the fraction of total trait variance that is heritable, i.e. due to inherited differences. In the infectious disease case, broad-sense heritability from pathogen factors, which are inherited by recipients from their infection partners, is typically measured. On the other hand, narrow-sense heritability $h^2$ measures the fraction of total trait variance due specifically to additive genetic effects, i.e. the sum of independent effects from all genetic variants. GWAS for infectious disease traits typically measure the narrow-sense heritability of a trait based on human genetic variants.

% How heritable is spVL (human side)
Several GWAS have been done to measure the narrow-sense heritability of spVL and identify associated host genetic variants \citep{Bartha2013, Dalmasso2008, Fellay2007, Pereyra2010, Fellay2009, Pelak2010, VanManen2011, McLaren2012}. The largest study to-date by \citet{McLaren2015} estimated the narrow-sense heritability of spVL from human genetic variants to be approximately 25\%. All but 5\% of this was attributed to two regions in the human genome, the major histocompatibility complex (MHC) and C-C motif chemokine receptor 5 (\emph{CCR5}). Both associations are biologically relevant: the MHC encodes proteins that present viral epitopes at the cell surface and \emph{CCR5} encodes a co-receptor for HIV-1 cell entry. In other words, MHC proteins match bits of the virus like puzzle pieces and display these to signal that a cell is infected. CCR5 proteins help the virus infect target cells.

% How heritable is spVL (viral side)
In addition to these human genetic factors, it is well-recognized that viral genetic factors affect spVL. As mentioned, heritability from the viral side is typically measured using broad-sense heritability. Estimates differ depending on the methods employed and the cohort studied (see \citet{Mitov2018} for a discussion of this uncertainty). Estimates using phylogenetic methods on large UK and Swiss cohorts by \citet{Mitov2018} and \citet{Bertels2018} measured the broad-sense heritability of spVL from the virus to be 21\% - 29\%. However, variation in the MHC is known to exert strong selective pressure on the virus \citep{Kloverpris2016, Nguyen2021SystematicLink}. If the virus can change its ``puzzle piece'' shape to escape MHC-presentation, infected cells can go undetected. This means that MHC variants affect spVL largely via selection on the virus \citep{Bartha2017}. In summary, human genetic factors play a role in determining spVL, but these effects may be due to interaction with specific viral genetic variants.

% The gap
Most of the GWAS for human genetic determinants of spVL conducted so far \citep{Dalmasso2008, Fellay2007, Pereyra2010, Fellay2009, Pelak2010, VanManen2011, McLaren2012, McLaren2015} do not explicitly consider any viral effect on spVL. In these GWAS, viral genetic effects are lumped in with residual variance due to other, non-genetic factors. This has several potential negative consequences. (i) Viral effects may be confounding or extraneous variables that bias estimates of host genetic effects. (ii) Variability due to viral effects would make it more challenging to identify human variants of small effect. Finally, (iii) spVL values from a cohort are not truly independent samples, given that patients closer in the transmission chain have more similar strains and therefore more similar spVL values.

@@

Our method is convenient for GWAS because it is simply a pre-processing step that produces an alternate response variable for GWAS association tests. It is still possible to use previously developed, well-documented, and fast tools for the actual association testing (we used PLINK \citep{Chang2015}). The method relies on the freely available R package POUMM \citep{Mitov2017a-POUMM} and all the code we wrote is available on the project GitHub at https://github.com/cevo-public/POUMM-GWAS. Future applications of our method might investigate other clinically significant disease traits and outcomes that are affected by both host and pathogen genetic factors, for instance Hepatitis B Virus-related hepatocellular carcinoma \citep{An2018HostCarcinoma}, Hepatitis C treatment success \citep{Ansari2017Genome-to-genomeVirus}, and susceptibility to or severity of certain bacterial infections, e.g. \cite{Messina2016ImpactInfections, Donnenberg2015BacterialStudies}.

@@

@@

We used the R package POUMM version 2.1.6 \citep{Mitov2017a-POUMM} to infer the POUMM parameters $g_0, \alpha, \theta, \sigma, \text{ and }\sigma_e$ from the approximate maximum-likelihood phylogeny and calculated spVL values. The Bayesian inference method implemented in this package requires specification of a prior distribution for each parameter. We used the same, broad prior distributions as in \citet{Mitov2018}, namely: $g_{0} \sim \mathcal{N}(4.5,\ 3)$, $\alpha \sim Exp(0.02)$, $\theta \sim \mathcal{N}(4.5,\ 3)$, $H^2_{\bar{t}} \sim \mathcal{U}(0,\ 1)$, and $\sigma^2_e \sim Exp(0.02)$. We ran two MCMC chains for 4x10$^6$ samples each with a target sample acceptance rate of 0.01 and a thinning interval of 1000. The first 2x10$^5$ samples of each chain were used for automatic adjustment of the MCMC proposal distribution. Figure \ref{fig:poumm-parameters} shows the posterior distributions for inferred parameters. Table \ref{tab:POUMM-parameters-our-estimates} gives the posterior mean values used for subsequent calculations.

\subsection*{Phylogenetic spVL correction}
We corrected calcualted spVL values using the method described in this paper. For each of the 1,392 individuals in the GWAS cohort, we estimated the viral part of spVL using equation \ref{eq:MuG} and the corresponding non-viral part using equation \ref{eq:EHat}. For the POUMM parameters $\alpha$, $\sigma$, $\theta$, and $\sigma_e$, we used the posterior mean estimates generated as described above. All the code used to implement the method is available at https://github.com/cevo-public/POUMM-GWAS.

@@

>>>>

\begin{abstract}
	Infectious diseases are particularly challenging for genome-wide association studies (GWAS) because genetic effects from two organisms (pathogen and host) may significantly influence a trait of interest. Previous GWAS have successfully identified host genetic variants associated with important infectious disease traits, such as HIV-1 set point viral load (spVL). However, these GWAS do not account for potentially confounding or extraneous pathogen effects that are heritable from donor to recipient in transmission chains.  We propose a new method to consider the full genome of each patient's infecting pathogen strain, remove strain-specific effects on a trait based on the pathogen phylogeny, and thus better estimate the effect of human genetic variants on infectious disease traits. In simulations, we show our method can increase GWAS power to detect truly associated host variants when pathogen effects are highly heritable, with strong phylogenetic correlations. When we apply our method to HIV-1 subtype B data from the Swiss HIV Cohort Study, we recover slightly weaker but qualitatively similar signals of association between spVL and human genetic variants in the \emph{CCR5} and major histocompatibility complex (MHC) gene regions compared to standard GWAS. Our simulation study confirms that based on the estimated heritability and selection parameters for HIV-1 subtype B spVL, standard GWAS are robust to pathogen effects. Our framework may improve GWAS for other diseases if pathogen effects are even more phylogenetically correlated amongst individuals in a cohort.
\end{abstract}

@@

% How do we measure the extent to which genetic factors determine spVL
To understand HIV pathogenicity, it is important to understand to what extent spVL is determined by host genetic factors as opposed to pathogen genetic factors. Heritability is a key measure of how genetically-determined a trait is. On the host side, narrow-sense heritability, the fraction of total trait variance explained by additive genetic effects, is a typical output of GWAS. 
% How heritable is spVL (human side)
Several GWAS have measured the narrow-sense heritability of spVL and identified associated host genetic variants \citep{Bartha2013, Dalmasso2008, Fellay2007, Pereyra2010, Fellay2009, Pelak2010, VanManen2011, McLaren2012}. The largest study to-date by \citet{McLaren2015} estimated that additive human genetic effects account for approximately 25\% of the variation in spVL amongst infected individuals. All but 5\% of this was attributed to two regions in the human genome, the major histocompatibility complex (MHC) and C-C motif chemokine receptor 5 (\emph{CCR5}). Both associations are biologically relevant: the MHC encodes proteins that present viral epitopes at the cell surface and \emph{CCR5} encodes a co-receptor for HIV-1 cell entry. In other words, MHC proteins match bits of the virus like puzzle pieces and display these to signal that a cell is infected. CCR5 proteins help the virus infect target cells.

% How heritable is spVL (viral side)
On the pathogen side, broad-sense heritability $H^2$ is typically measured. This is the fraction of total trait variance due to any factors ``inherited'' from infection partner to infection partner (the most obvious being pathogen genetic factors). Broad-sense heritability is commonly measured by quantifying how correlated a trait is between infection partners or how phylogenetically structured a trait is based on a reconstructed pathogen phylogeny. For HIV spVL, estimates differ depending on the methods employed and the cohort studied (see \citet{Mitov2018} for a discussion of this uncertainty). Using phylogenetic methods on large UK and Swiss cohorts, \citet{Mitov2018} and \citet{Bertels2018} estimated that pathogen factors account for 21\% - 29\% of the variation in spVL amongst infected individuals. However, variation in the MHC is known to exert strong selective pressure on the virus \citep{Kloverpris2016, Nguyen2021SystematicLink}. If the virus can change its ``puzzle piece'' shape to escape MHC-presentation, infected cells can go undetected. This means that MHC variants affect spVL largely via selection on the virus \citep{Bartha2017}. In summary, human genetic factors play a role in determining spVL, but these effects may be due to interaction with specific viral genetic variants.

% The gap
Most of the GWAS for human genetic determinants of spVL conducted so far \citep{Dalmasso2008, Fellay2007, Pereyra2010, Fellay2009, Pelak2010, VanManen2011, McLaren2012, McLaren2015} do not explicitly consider any viral effect on spVL. In these GWAS, viral genetic effects are lumped in with residual variance due to other, non-genetic factors. This has several potential negative consequences. Viral effects may be (i) confounding or (ii) extraneous variables that produce distorted or spurious associations between host genetic variants and a disease trait in host GWAS. 
%This may occur if the virus affects the disease trait (the dependent variable in host GWAS) or if it affects both the disease trait and the host genotype (the independent variable in host GWAS).
For example, stratification in both the host and pathogen populations may yield spurious correlations in host GWAS that are really due to independent pathogen factors. Then, co-evolution between a pathogen and its host may yield distorted estimates for host genetic effects that are not broadly protective against all pathogen strains. Finally, (iii) spVL values from a cohort are not truly independent samples, given that patients closer in the transmission chain have more similar strains and therefore more similar spVL values.

@@

Our method is convenient for GWAS because it is simply a pre-processing step that produces an alternate response variable for GWAS association tests. It is still possible to use previously developed, well-documented, and fast tools for the actual association testing (we used PLINK \citep{Chang2015}). The method relies on the freely available R package POUMM \citep{Mitov2017a-POUMM}, which scales to trees of up to 10,000 tips \citep{Mitov2019}. All code for the simulations and HIV spVL analysis presented in this study is available on the project GitHub at https://github.com/cevo-public/POUMM-GWAS. Future applications of our method might investigate other clinically significant disease traits and outcomes that are affected by both host and pathogen genetic factors, for instance Hepatitis B Virus-related hepatocellular carcinoma \citep{An2018HostCarcinoma}, Hepatitis C treatment success \citep{Ansari2017Genome-to-genomeVirus}, and susceptibility to or severity of certain bacterial infections, e.g. \cite{Messina2016ImpactInfections, Donnenberg2015BacterialStudies}.

@@

\subsection*{\emph{A. thaliana}-\emph{X. arboricola} data}
\emph{A. thaliana} genotypes were downloaded from the 1001 Genomes project \citep{Alonso-Blanco20161135Thaliana}. Quantitative diseases resistance (QDR) score measurements and \emph{X. arboricola} genotypes were provided by the authors of \citep{Wang2018Two-wayGenomes}. The \emph{X. arboricola} QDR trait values were generated in a fully-crossed experimental design, where each host type (accession) was infected with each pathogen strain. QDR was measured on 4 leaves for each host-pathogen pairing.

For phylogenetic inference, the \emph{X. arboricola} genotype matrix consisted of allele dosage information for 33,610 SNPs in 22 strains. We used the ``dist.gene'' and ``nj'' functions from the ape package in R to construct a pairwise genetic distance matrix and then a neighbor-joining tree from these data. We calculated mean QDR across all leaves infected on all hosts for each \emph{X. arboricola} strain.

For GWAS, we retained data from 133 \emph{A. thaliana} accessions with associated QDR measurements.

@@

We used the R package POUMM version 2.1.6 \citep{Mitov2017a-POUMM} to infer the POUMM parameters $g_0, \alpha, \theta, \sigma, \text{ and }\sigma_e$ from the HIV-1 and \emph{X. arboricola} phylogenies and associated spVL and QDR trait values. The Bayesian inference method implemented in this package requires specification of a prior distribution for each parameter. For HIV-1 spVL, we used the same, broad prior distributions as in \citet{Mitov2018}, namely: $g_{0} \sim \mathcal{N}(4.5,\ 3)$, $\alpha \sim Exp(0.02)$, $\theta \sim \mathcal{N}(4.5,\ 3)$, $H^2_{\bar{t}} \sim \mathcal{U}(0,\ 1)$, and $\sigma^2_e \sim Exp(0.02)$. For \emph{X. arboricola} QDR, we modified the $g_0$ and $\theta$ priors to match the empirical mean and standard deviation of QDR trait values in the dataset: $g_{0} \sim \mathcal{N}(0.4,\ 0.2)$ and $\theta \sim \mathcal{N}(0.4,\ 0.2)$. We ran two MCMC chains for 4x10$^6$ samples each with a target sample acceptance rate of 0.01 and a thinning interval of 1000 for both analyses. The first 2x10$^5$ samples of each chain were used for automatic adjustment of the MCMC proposal distribution. Figure \ref{fig:poumm-parameters} shows the posterior distributions for inferred parameters. Table \ref{tab:POUMM-parameters-our-estimates} gives the posterior mean values used for subsequent calculations.

\subsection*{Phylogenetic trait correction}
We corrected HIV-1 spVL and \emph{X. arboricola} QDR trait values using the method described in this paper. For each host individual, we estimated the pathogen part of their trait value using equation \ref{eq:MuG} and the corresponding non-pathogen part using equation \ref{eq:EHat}. For the POUMM parameters $\alpha$, $\sigma$, $\theta$, and $\sigma_e$, we used the posterior mean estimates generated as described above. For the \emph{A. thaliana}-\emph{X. arboricola} trait correction and downstream GWAS, we randomly selected one pathogen pairing for each host to analyze. This makes the dataset analogous to the observational HIV-1 dataset. All the code used to implement the method is available at https://github.com/cevo-public/POUMM-GWAS.

@@

\begin{table}[H]
\caption{Number of variants for \emph{X. arboricola} QDR GWAS after sequential filtering steps.}
	\begin{tabularx}{\linewidth}{XX}
	\hline
	Variant filter & Number of variants remaining \\
	\hline 
	Raw data & 12883854 \\
	Bi-allelic variants & 11769920 \\ \hline			
	\end{tabularx}
	\label{tab:variant-filtering-plants}
\end{table}
